Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2015-02-05 doc/ttsp2013/paper/ttsp_2013.tex
updates to ttsp data

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to computing a solution over the entire
domain a reactor core where diffusion theory is often valid.

@@

To enable solutions of the $SP_N$ equations for these problems, this
work develops two new techniques. First, we define an algebraic
preconditioning strategy for MCSA that leverages an explicitly formed
preconditioner as well as a matrix-matrix product to construct the
matrix elements of the preconditioned system. As a preconditioner,
this work uses a sparse approximate inverse technique which results in
a sparse matrix with element-wise access, a necessary requirement for
implementation of the Neumann-Ulam Monte Carlo algorithm. Second, we
develop a new domain decomposition scheme for MCSA with a new fully
asynchronous domain decomposed Monte Carlo kernel for the Neumann-Ulam
algorithm. In tandem, these two developments permit solutions of the
$SP_N$ equations for large-scale light water criticality problems. We
believe both of these contributions in preconditioning and
parallelizing the Neumann-Ulam algorithm are new and unique.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work. We then describe the application of the MCSA
algorithm to the solution of the equations and introduce the
criticality problems used for the numerical studies in the this
work. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for these problems and subsquently present our new preconditioning
scheme. We then parallelize the MCSA algorithm and describe our new
domain decomposition scheme deisgned to enable very large problems. We
then apply this preconditioning scheme and parallel algorithm to the
criticality problems and compare the performance of the preconditioned
MCSA method to preconditioned Krylov methods. We then finish with
conclusions and indication of future work.

@@

Convergence of the Monte Carlo method, previously assumed by many to
be entirely bound by the spectral radius of $\ve{H}^T$
(e.g. \cite{spanier_monte_1969}), was recently identified as a only a
subset of the convergence criteria. In particular, Ji, Mascagni, and
Li demonstrate in \cite{ji_2013} that a necessary and sufficient
condition for convergence of this algorithm is $\rho(\ve{H^*}) < 1$
with $\ve{H^*}$ defined element-wise as:
\begin{equation}
  h^*_{ij} = \frac{(h^T_{ij})^2}{p_{ij}}\:,
  \label{eq:h_star}
\end{equation}
which is indicative that both the structure of $\ve{H}^T$ and the
selection of transition probabilities affect the convergence of the
algorithm.

@@

%%---------------------------------------------------------------------------%%
\section{Preconditioning Algorithm}

\subsection{Preconditioned MCSA}
\label{subsec:preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this by altering the eigenvalue spectrum of the
iteration matrix. It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underlying
Monte Carlo problem that will be solved with the Neumann-Ulam
method. We consider here the general left/right preconditioned method
as the left or right preconditioned methods can be inferred from its
formulation.

We consider a left preconditioner $\ve{M_L}$ and a right
preconditioner $\ve{M_R}$. The left/right preconditioned linear
problem is then:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
  \label{eq:left_right_linear_problem}
\end{equation}
To handle the right preconditioning, the system is written with a
substitution of variables:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
  \label{eq:left_right_subs_problem}
\end{equation}
with
\begin{equation}
  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
  \label{eq:left_right_recover}
\end{equation}
To apply such a method to MCSA, we solve for the substituted variable
$\ve{u}$ during the iteration sequence:
\begin{subequations}
  \begin{gather}
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
  \end{gather}
  \label{eq:left_right_mcsa}
\end{subequations}
and then recover the original solution vector with
Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
isolate the generation of the correction:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
  \label{eq:left_right_correction}
\end{equation}
and note that the preconditioned residual of the substituted variable
is now serving as the source and the new iteration matrix is:
\begin{equation}
  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
  \label{eq:left_right_iteration_matrix}
\end{equation}
As we require $(i,j)$ element-wise access to the iteration matrix in
order to construct probabilities and weights for the Monte Carlo
procedure from the Neumann-Ulam decomposition, the \textit{composite
  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
matrix-matrix multiplication.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
part of a sequence for finding the substituted variable $\ve{u}$. We
do, however, gain the benefits of this general preconditioning by
building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator.

\subsubsection{Block Jacobi Preconditioning}

\subsubsection{Sparse Approximate Inverse Preconditioning}

>>>>

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware, and
modern algorithms. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to computing a solution over the entire
domain of a reactor core where diffusion theory is often valid.

@@

To enable solutions of the $SP_N$ equations for these problems, this
work develops two new techniques. First, we define an algebraic
preconditioning strategy for MCSA that leverages an explicitly formed
preconditioner as well as a matrix-matrix product to construct the
matrix elements of the preconditioned system. As a preconditioner,
this work uses a sparse approximate inverse technique which results in
a sparse matrix with element-wise access, a necessary requirement for
implementation of the Neumann-Ulam Monte Carlo algorithm. Second, we
develop a new domain decomposition scheme for MCSA with a new fully
asynchronous domain decomposed Monte Carlo kernel for the Neumann-Ulam
algorithm. In tandem, these two developments permit solutions of the
$SP_N$ equations for large-scale light water criticality problems. We
believe both of these contributions in preconditioning and
parallelizing the Neumann-Ulam algorithm are new and unique. Although
we have motivated this work by looking forward to exascale computing,
resiliency of this solution technique in the presence of faults will
not be considered here but instead be addressed in future work.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work. We then describe the application of the MCSA
algorithm to the solution of the equations and introduce the
criticality problems used for the numerical studies in the this
work. Next, we present our new preconditioning scheme consisting of an
approximate inverse algorithm combined with solving an approximate
Monte Carlo problem for the acceleration. We then parallelize the MCSA
algorithm and describe our new domain decomposition scheme for the
Neumann-Ulam Monte Carlo algorithm designed to enable very large
problems. We then apply the preconditioning scheme and the parallel
algorithm to the criticality problems and compare the performance of
the preconditioned MCSA method to preconditioned Krylov methods. We
then finish with conclusions and indication of future work.

@@

Convergence of the Monte Carlo method, previously assumed by many to
be entirely bound by the spectral radius of $\ve{H}^T$
(e.g. \cite{spanier_monte_1969}), was recently identified as a only a
subset of the convergence criteria. In particular, Ji, Mascagni, and
Li demonstrate in \cite{ji_2013} that a necessary and sufficient
condition for convergence of this algorithm is 
\begin{equation}
  \rho(\ve{H^*}) < 1
  \label{eq:necessary_condition}
\end{equation} 
with $\ve{H^*}$ defined element-wise as:
\begin{equation}
  h^*_{ij} = \frac{(h^T_{ij})^2}{p_{ij}}\:,
  \label{eq:condition_values}
\end{equation}
or for our choice of probabilities via
Eq~(\ref{eq:adjoint_neumann_ulam})
\begin{equation}
  h^*_{ij} = h_{ij} w_{ij}\:,
  \label{eq:h_star}
\end{equation}
which is indicative that both the structure of $\ve{H}^T$ and the
selection of transition probabilities affect the convergence of the
algorithm.

@@

%%---------------------------------------------------------------------------%%
\section{Preconditioning Algorithm}
In this section we present our preconditioning algorithm.

\subsection{Preconditioned MCSA}
\label{subsec:preconditioning}
In most cases, at least a minimal amount of preconditioning of the
linear system will be required in order to satisfy the necessary
convergence condition given in Eq~(\ref{eq:necessary_condition}) for
the Neumann-Ulam algorithm. Preconditioning serves as a means of
achieving this by altering the eigenvalue spectrum of the adjoint
iteration matrix and subsequently $\ve{H}^*$. It is possible to use
general left and right preconditioning with MCSA by carefully
considering the underlying Monte Carlo problem that will be solved
with the Neumann-Ulam algorithm. We consider a left preconditioner,
$\ve{M_L}$, and a right preconditioner, $\ve{M_R}$. The left/right
preconditioned linear problem is then:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
  \label{eq:left_right_linear_problem}
\end{equation}
To handle the right preconditioning, the system is written with a
substitution of variables:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
  \label{eq:left_right_subs_problem}
\end{equation}
with
\begin{equation}
  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
  \label{eq:left_right_recover}
\end{equation}
To apply this preconditioning to MCSA, we solve for the substituted
variable $\ve{u}$ during the iteration sequence:
\begin{subequations}
  \begin{gather}
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{M}_L^{-1}\ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2}\:,\\ 
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{M}_L^{-1}\ve{r}^{k+1/2}\:,\\ 
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1}\:,
  \end{gather}
  \label{eq:left_right_mcsa}
\end{subequations}
and then recover the original solution vector with
Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
isolate the generation of the correction:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} =
  \ve{M}_L^{-1}\ve{r}^{k+1/2}\:,
  \label{eq:left_right_correction}
\end{equation}
and note that the preconditioned residual of the substituted variable
is now serving as the source and the new adjoint iteration matrix is:
\begin{equation}
  \ve{H}^T = \big(\ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\big)^T\:.
  \label{eq:left_right_iteration_matrix}
\end{equation}
As we require element-wise access to the iteration matrix in order to
construct probabilities and weights from the Neumann-Ulam
decomposition for the Monte Carlo procedure, the composite operator,
$\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via matrix-matrix
multiplication. Provided that $\ve{M}_L^{-1}$, $\ve{A}$, and
$\ve{M}_R^{-1}$ are relatively sparse, then we can readily leverage
sparse matrix-matrix multiplication algorithms that are in wide use in
scalable algebraic multigrid implementations.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
is part of a sequence for finding the substituted variable,
$\ve{u}$. We do, however, gain the benefits of this general
preconditioning by building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator.

\subsection{Sparse Approximate Inverse Preconditioning}

\subsection{Approximate Monte Carlo Problem}
